---
title: "Lab 4"
author: "Shan He, Joanna Huang, Tiffany Jaya"
date: "18 December 2017"
output: pdf_document
---

TODO: As this is a policy exercise, you should do your best to address the campaign’s questions from a causal perspective. At the same time, you should clearly explain the limitations of your analysis, and provide discussion around whether your estimates suffer from endogeneity bias.

# Introduction 

TODO: A brief introduction

# Exploratory Data Analysis

TODO: An initial exploratory analysis. Detect any anomalies, including missing values, top-coded or bottom- coded variables, etc.


```{r}
library(car) # lm
library(ggplot2) # ggplot
library(lmtest) # bptest
library(plm) # vcovHC
library(sandwich) 
library(stargazer) # stargazer
library(tidyr) # gather
```

```{r}
data <- read.csv("crime.csv")
summary(data)
```

sample size n = 90
num of var = 26
Summary ranges for variables look reasonable except for prbarr (‘probability’ of arrest) and prbconv (‘probability’ of conviction) where there appears to be some outliers that have a probability greater than 1.0.

```{r}
colSums(sapply(data, is.na))
```

There are no missing values. 

```{r}
plot.data <- na.omit(data[, sapply(data, is.numeric)])
ggplot(gather(plot.data), aes(value)) + 
       facet_wrap(~key, scales="free") + 
       geom_histogram()
```
# Observations
- All observations were recorded for the year of 1987. 
- $urban$, $central$, $west$ are categorical variables.
- Most wages variables (wcon, wtrd, wfir, wser, wmfg, wfed, wsta) appear to be slightly positively skewed which may be due to less people getting paid towards the higher end. 
- wtuc, wloc appear to be normally distributed.
- prbarr, prbconv appear to be positively skewed while prbpris is more negatively skewed.
- avgsen appears to be positively skewed.
- crmrte, density, polpc, prbconv: The histogram indicates that these variables have positive skews. Given the variables have a meaningful zero-point, we can take the log for a more normal distribution.
- Question: what is mix? 

```{r}
plot.data <- na.omit(log(data[, sapply(data, is.numeric)]))
ggplot(gather(plot.data), aes(value)) + 
       facet_wrap(~key, scales="free") + 
       geom_histogram()
```
# Observations after taking logs:
It looks like the log transformation of variables cmrte, mix, prbconv, wfed, wfir, wmfg, and wsta have made the distributions quite normal. This will help ensure the errors of the model are normal.

# Proposed Model
Based on the exploratory data analysis and general intution about the potential determinants of crime, an initial proposed model specification and coefficient expectation are:

- Sum of all wages: Summing the wage variables across sectors may identify whether income inequality between counties may explain the difference in crime rates. Initial thoughts are that sum of wages (higher incomes) lead to lower crime rates, so a linear relationship between sum of wages and crimes committed per person is expected with a negative coefficient.

- log(prbarr) Probability of arrest: There is the possibility that the probability of arrest is positively correlated with crime rate, as higher numbers of arrest can increase the number of crimes recorded.

- tax revenue per capita (taxpc): Lack of government-funded resources can take the form of a lack of educational opportunities or employment options, thus leading to rise in crime rates.

Model 1:
$log(crmrte) = B_0 + B_1(allwages) + B_2log(prbarr) + B_3(taxrev)+u$


# Model Building Process

TODO: A model building process, supported by exploratory analysis. Your EDA should be interspersed with, and support, your modeling decisions. In particular, you should use exploratory techniques to address
* What transformations to apply to variables and what new variables should be created.
* What variables should be included in each model
* Whether model assumptions are met

Potential independent variables: polpc (police per capita), density (people per sq mile), taxpc (tax revenue per capita), pctmin80 (percentage minority in 1980), pctymle (percentage young male)

```{r}
pairs(crmrte ~ polpc + density + taxpc + pctmin80 + pctymle, data=data)
```

```{r}
# number of urban
sum(data$urban == 1)
# number of central
sum(data$central == 1)
# number of west
sum(data$west == 1)
```



# Model Specifications

TODO: 
* One model with only the explanatory variables of key interest (possibly transformed, as determined by your EDA), and no other covariates.
* One model that includes key explanatory variables and only covariates that you believe increase the accuracy of your results without introducing bias (for example, you should not include outcome variables that will absorb some of the causal effect you are interested in). This model should strike a balance between accuracy and parsimony and reflect your best understanding of the determinants of crime.
* One model that includes the previous covariates, and most, if not all, other covariates. A key purpose of this model is to demonstrate the robustness of your results to model specification.

For your first model, a detailed assessment of the 6 CLM assumptions. For additional models, you should check all assumptions, but only highlight major differences from your first model in your report.

A well-formatted regression table summarizing your model results. Make sure that standard errors presented in this table are valid. Also, be sure to comment on both statistical and practical significance.

# Causality

TODO: A detailed discussion of causality. In particular, include a discussion of what variables are not included in your analysis and the likely direction of omitted variable bias. Highlight any coefficients you find that appear to have the wrong sign from a causal perspective, and explain why this is the case.

# Conclusion

TODO: A brief conclusion with a few high-level takeaways.
