---
title: "Lab 4"
author: "Shan He, Joanna Huang, Tiffany Jaya"
date: "17 December 2017"
output: pdf_document
---

```{r echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(car) # scatterplotMatrix
library(ggplot2) # ggplot
library(moments) # skewness
library(outliers) # outlier
library(tidyr) # gather
# prevent scientific notation, specifically used in outliers package
options(scipen=999) 
```

# Introduction

The purpose of this report is to generate policy suggestions based on our understanding of the determinants of crime in North Carolina in 1987. We will list out the limitations of our analysis, including any estimates that suffer from endogeneity bias. 

# Exploratory Data Analysis

```{r}
# load the data
data <- read.csv("crime.csv")
# verify that it only contains data from 1987
unique(data$year)
# list number of counties
length(unique(data$county))
# list number of western, central, and urban counties
c(sum(data$west == 1), sum(data$central == 1), sum(data$urban == 1))
# list number of western & urban counties and central & urban counties
c(sum(data$west == 1 & data$urban == 1), sum(data$central == 1 & data$urban == 1))
# verify number of missing values
colSums(sapply(data, is.na))
```

The dataset contains 90 counties from North Carolina, all of which is collected in 1987. Out of the 90 counties, 21 are from western NC (out of which 1 is also urban), 34 are from central NC (out of which 5 is also urban), and 8 are considered urban counties. There are no missing values which will make our analysis easier. 

For now, we will not take into consideration probabilities that are greater than 1 or less than 0 as well as percentages that are greater than 1 or less than 0. The assumption is that probabilities are in the range [0, 1] and percentages are in the range [0, 100]. Until we know the reason why the values are outside their range, we will not employ datapoints that do not conform to this assumption.

```{r}
# list number of probabilities (prbarr, prbconv, prbpris, mix) that are not in range [0, 1]
c(sum(data$prbarr < 0 | 1 < data$prbarr), sum(data$prbconv < 0 | 1 < data$prbconv), 
  sum(data$prbpris < 0 | 1 < data$prbpris), sum(data$mix < 0 | 1 < data$mix))
# list number of percentages (pctymle, pctmin80) that are not in range [0, 100]
c(sum(data$pctymle < 0 | 100 < data$pctymle), sum(data$pctmin80 < 0 | 100 < data$pctmin80))
```

$prbarr$ and $prbconv$ contain 1 and 10 datapoints respectively that do not conform to the probability assumption.

We then plot each numeric variable in a histogram to see its sample distribution. 

```{r message=FALSE}
# plot every variable except X, county, year, west, central, urban
num.data <- data[!(names(data) %in% c("X", "county", "year", "west", "central", "urban"))]
ggplot(gather(num.data), aes(value)) + 
       facet_wrap(~key, scales="free") + 
       geom_histogram()
```

```{r}
skewness(num.data)
```

Most of the sample distributions appear to be positively skewed. We will take into consideration of logarithmic transformations, depending whether the interpretations make sense, when it is time to include the variables into the regression model.

From the histograms, we also see several notable outliers. We are under the impression that a county which has outlier in one variable will likely have outlier in another variable. For this reason, we have listed counties which have repeated outliers when we iterate through the entire numeric variables. 

```{r}
# iterate through each numeric variable and list the outlier counties and their respective frequency 
county.ids <- c()
for(var in num.data) {
  var.out <- boxplot.stats(var)$out
  county.ids <- c(county.ids, data[var %in% var.out, ]$county)
}
table(county.ids)
```
```{r}
# list the most extreme outlier
outlier(num.data)
```

One outlier that is interesting to note is that the weekly wage in the service industry for county with id 185 is \$2177.10, which is approximately eight times higher than the median. We do not know if the value is inputted incorrectly or if the county in general is making a weekly wage of \$2177.10 in the service industry.

```{r}
summary(data$wser)
```

# Research Question

James Q. Wilson and George Kelling's "broken windows theory" in 1982 led to a nation-wide movement for stricter crime-fighting policies between the 1980s and 1990s. The theory states:  

> *if the first broken window in a building is not repaired, then people*
> *who like breaking windows will assume that no one cares about the* 
> *building and more windows will be broken. Soon the building will have* 
> *no windows....* 

The belief was that by adopting a zero tolerance approach that enforced even the lowest level offenses, crime rates would subsequently go down. While New York City notably enforced this more stringent approach, San Francisco went the opposite direction of less strident law enforcement policies that reduced arrests, prosecutions and incarceration rates. Both sides experienced considerable declines in crime rates. Thus we hope to test the "broken windows theory" for the counties of South Carolina in 1987 and answer the question: Does the conservative approach of deterrence through arrests, incapacitation through imprisonment, harsh sentencing and higher police per capita lead to lower crime rates? 

# Model 1: only the explanatory variables of key interest

Based on the exploratory data analysis and goal of testing the "broken windows theory", our initial proposed model will include all variables related to stricter law enforcement policies: *prbarr*, *prbconv*, *prbpris*, *avgsen*, and *polpc*. Assuming the "broken windows theory" is valid, we expect generally negative coefficients for all variables.

To make *prbarr*, *prbconv* and *prbpris* interpretation more intuitive to understand, we decided to transform them into percentages, and will use these new variables moving forward.
```{r}
data$arrperc = data$prbarr * 100
data$convperc = data$prbconv * 100
data$prisperc = data$prbpris * 100
```

In our EDA, we noted that crmrte, prbarr, prbconv, avgsen, and polpc are positively skewed, thus we should consider taking the log. Since the values cannot be negative and have meaningful zero-points, this decision is valid. It is important to remember moving forward that the interpretation of arrperc, and convperc will be in terms of percentage point changes rather than percentage changes.

```{r}
plot.data <- na.omit(data[, sapply(data, is.numeric)])
 ggplot(gather(plot.data), aes(value)) + 
        facet_wrap(~key, scales="free") + 
        geom_histogram()
plot.data <- na.omit(log(data[, sapply(data, is.numeric)]))
 ggplot(gather(plot.data), aes(value)) + 
        facet_wrap(~key, scales="free") + 
        geom_histogram()
 
```

```{r}
scatterplotMatrix(~log(crmrte) + log(arrperc) + log(convperc) + log(polpc) + log(avgsen) + prisperc, data = data)
 + cor(data[, c("crmrte", "arrperc", "convperc", "polpc", "avgsen", "prisperc")])
 + cor(data[!(names(data) %in% c("X", "county", "year"))], use = "complete.obs")
```
Looking at the relationships between crmrte (y) and the newly transformed x variables, it appears that the relationships are linear and we can continue with our proposed multiregression model.

The model:

$$ log(crmrte) = \beta_0 + \beta_1log(arrperc) + \beta_2log(convperc) + \beta_3(prisperc) + \beta_4log(avgsen) + \beta_5log(polpc) + \mu $$

We will now run the model and test the validity of the 6 CLM assumptions:
```{r}
m1 = lm(log(crmrte) ~ log(arrperc) + log(convperc) + prisperc + log(avgsen) + log(polpc), data=data)

```


## CLM 1 - A linear model

The model is specified such that the dependent variable is a linear function of the explanatory variables. 

Is the assumption valid? **Yes**

**Response**: No response required.

## CLM 2 - Random Sampling

Is the assumption valid? 

**Response**: 

## CLM 3 - Multicollinearity

As a quick test of the multicollinearity condition, we check the correlation of the two explanatory variables and their Variance Inflation Factors (VIF):
```{r}
data$logarrperc = log(data$arrperc)
data$logconvperc = log(data$convperc)
data$logpolpc = log(data$polpc)
data$logavgsen = log(data$avgsen)
X = data.matrix(subset(data, select=c("logarrperc", "logconvperc", "logpolpc", "logavgsen", "prisperc")))
(Cor = cor(X))
```
```{r}
vif(m1)
```

The explanatory variables (logarrperc, logconvperc, logpolpc, logavgsen and prisperc) are not perfectly correlated and the VIFs are low (i.e. less than 10), so there is no perfect multicollinearity of the independent variables.

Is the assumption valid? **Yes** 
**Response**: No response required.

## CLM 4 – Zero-Conditional Mean
To see whether there is a zero-conditional mean across all x’s, we will plot the residuals against the fitted values.

```{r}
plot(m1, which=1)
```

```{r}
plot(m1, which=5)
```

Looking at the residuals vs. leverage plot, it appears there are a couple outliers that have considerable leverage on the regression, one of which also has a Cook's distance that's greater than 1. 

```{r}
data[data$X == 51 | data$X == 79,]
```
Upon closer inspection, the outlier, record 51, has percentages above 100% for both arrest and conviction. Since this is not actually possible, we should consider removing this outlier.

```{r}
(cov(data$logarrperc,m1$residuals))
(cov(data$logconvperc,m1$residuals))
(cov(data$logavgsen,m1$residuals))
(cov(data$logpolpc,m1$residuals))
(cov(data$prisperc,m1$residuals))
```

The plots indicates little evidence that the zero-conditional mean assumption doesn’t hold, as the red spline line remains close to zero despite its slight dip and rise at both ends due to less observations.

The covariances of the three independent variables with the residuals are very close to zero indicating they are likely exogenous.

One data point has a large Cook’s distance and may have undue influence on the model fit.

Is the assumption valid? 

**Response**: 

## CLM 5 - Homoscedasticity
To determine whether the variance of $\mu$ is fixed for all x’s, we will first take a look at the residuals plotted against the fitted values to see whether the variance of residuals is constant across the fitted values.  
```{r}
plot(m1, which=1)
```
The plot indicates no strong evidence of heteroskedasticity. 

To further understand whether the model meets homoskedasticity, we will perform statistical tests Breusch-Pagan and the Score-test for non-constant error variance.
```{r}
library(lmtest)
bptest(m1)
```
With a p-value of 0.049, we cannot reject the null hypothesis of homoskedasticity at the 5% significance level.

```{r}
ncvTest(m1)
```

With a p-value of .057, we cannot reject the hypothesis of constant error variance.

Is the assumption valid? **Yes**
**Response**: Though we can assume homoskedasticity, we will move forward with robust standard errors, as they provide more accurate p-values.

##CLM 6 – Normality of residuals
To determine whether there is normality of the residuals, we will use a histogram or Q-Q plots of the residuals and visually observe whether there is normality.

```{r}
# normality of standard residuals
rstnd = rstandard(m1)
hist(rstnd, main="Histogram standard residuals", breaks = 50, freq=FALSE)
curve(dnorm(x, mean=0, sd=sd(rstnd)), col="red", lwd=2, add=TRUE)
```
```{r}
# normality of studentized residuals
stud = rstudent(m1)
hist(stud, main="Histogram studentized residuals", breaks = 50, freq=FALSE)
curve(dnorm(x, mean=0, sd=1), col="red", lwd=2, add=TRUE)
```

```{r}
# Q-Q plot standard residuals
qqPlot(rstnd, distribution="norm", pch=20, main="QQ-Plot standard residuals")
qqline(rstnd, col="red", lwd=2)
```

```{r}
# Q-Q plot studentized residuals
qqPlot(stud, distribution="norm", pch=20, main="QQ-Plot studentized residuals")
qqline(stud, col="red", lwd=2)
```
The histograms of the residuals have a negative skew, and QQ-plots further demonstrate nonormality in the error distribution.

Is the assumption valid? **No**
Response: 

# Model 2: add covariates that increase accuracy without bias

crmrte ~ avgsen + prbarr + prbconv + prbpris + polpc

```{r}
hist(data$avgsen)
hist(data$prbarr)
hist(data$prbconv)
hist(data$prbpris)
hist(data$polpc)
```

# Model 3: most, if not all, other covariates

Adding arbitrary variables uses up valuable degrees of freedom which increases the standard error which in turn reduces the precision of this estimate. This effect is especially pronounced in small sample size. 

# Summary of Models

# Discussion of Causality

# Conclusion

# References:
"Shattering "Broken Windows": An Analysis of San Francisco’s Alternative Crime Policies", CENTER ON JUVENILE AND CRIMINAL JUSTICE, October 1999 http://www.cjcj.org/uploads/cjcj/documents/shattering.pdf
