---
title: "Lab 4"
author: "Shan He, Joanna Huang, Tiffany Jaya"
date: "18 December 2017"
output: pdf_document
---

# Introduction

The purpose of this report is to generate policy suggestions based on our understanding of the determinants of crime in North Carolina in 1987. We will list out the limitations of our analysis, including any estimates that suffer from endogeneity bias. 

# Exploratory Data Analysis

```{r}
# load the data
data <- read.csv("crime.csv")
# verify that it only contains data from 1987
unique(data$year)
# list number of counties
c(nrow(data))
# list number of western, central, and urban counties
c(sum(data$west == 1), sum(data$central == 1), sum(data$urban == 1))
# list number of western & urban counties and central & urban counties
c(sum(data$west == 1 & data$urban == 1), sum(data$central == 1 & data$urban == 1))
# verify number of missing values
colSums(sapply(data, is.na))
```

The dataset contains 90 counties from North Carolina, all of which is collected in 1987. Out of the 90 counties, 21 are from western NC (out of which 1 is also urban), 34 are from central NC (out of which 5 is also urban), and 8 are considered urban counties. There are no missing values which will make our analysis easier. 

For now, we will not take into consideration probabilities that are greater than 1 or less than 0 as well as percentages that are greater than 1 or less than 0. The assumption is that probabilities are in the range [0, 1] and percentages are in the range [0, 100]. Until we know the reason why the values are outside their range, we will not employ datapoints that do not conform to this assumption.

```{r}
# list number of probabilities (prbarr, prbconv, prbpris, mix) that are not in range [0, 1]
c(sum(data$prbarr < 0 | 1 < data$prbarr), sum(data$prbconv < 0 | 1 < data$prbconv), sum(data$prbpris < 0 | 1 < data$prbpris), sum(data$mix < 0 | 1 < data$mix))
# list number of percentages (pctymle, pctmin80) that are not in range [0, 100]
c(sum(data$pctymle < 0 | 100 < data$pctymle), sum(data$pctmin80 < 0 | 100 < data$pctmin80))
```

$prbarr$ and $prbconv$ contain 1 and 10 datapoints respectively that do not conform to the probability assumption.

We have also decided to create an additional column that adds up all the wages to see if wages collectively can be considered as a predictor variable for the regression.

```{r}
# create a column that adds up all the wages
data$wtotal <- rowSums(subset(data, select=c("wcon", "wtuc", "wtrd", "wfir", "wser", "wmfg", "wfed", "wsta", "wloc")))
```

We then plot each numeric variable in a histogram to see its sample distribution. 

```{r}
# plot every variable except X, county, year, west, central, urban
library(ggplot2) # ggplot
library(tidyr) # gather
num.data <- data[!(names(data) %in% c("X", "county", "year", "west", "central", "urban"))]
ggplot(gather(num.data), aes(value)) + 
       facet_wrap(~key, scales="free") + 
       geom_histogram()
```

```{r}
library(moments) # skewness
skewness(num.data)
```

Most of the sample distributions appear to be positively skewed. We will take into consideration a logarithmic transformation when it is time to include the variables into the regression model.

From the histograms, we also see several notable outliers. We are under the impression that a county which has outlier in one variable will have outlier in another variable. For this reason, we have listed counties which have repeated outliers when we iterate through the entire numeric variables. 

```{r}
# iterate through each numeric variable and list the outlier counties and their respective frequency 
county.ids <- c()
for(var in num.data) {
  var.out <- boxplot.stats(var)$out
  county.ids <- c(county.ids, data[var %in% var.out, ]$county)
}
table(county.ids)
```

One outlier that is interesting to note is that the weekly wage in the service industry for county with id 185 is \$2177.10, which is approximately eight times higher than the median. We do not know if the value is inputted incorrectly or if the county in general is making a weekly wage of \$2177.10 in the service industry.

```{r}
summary(data$wser)
```

## Variable Selection 

```{r}
# perform stepwise selection both forward and backward
library(MASS)
#m.all <- lm(crmrte ~ )
```

```{r}
# perform all-subsets regression
library(leaps)
#leaps <- regsubsets(crmrte ~ )
```

